---
title: <center>Ontogenetic Migration and Habitat Selection by Invasive Lionfish</center>
author: <center>Alisha Naidu^1^, Juan Carlos Villaseñor-Derbez^2^  </center>
date: <center>^1^*Lynbrook High School, 1280 Johnson Avenue, San Jose, CA 95129*  </center> <center>*alisha415naidu@gmail.com*  </center> <center>^2^*Bren School of Environmental Science and Management, University of California, Santa Barbara, CA 93106*  </center>
abstract: "The introduction of invasive lionfish into the Western Atlantic has greatly reduced local fish populations through predation, representing a major threat to native organisms and ecosystem managers alike. Although observational evidence suggests that lionfish size increases with depths, this pattern has not been formally addressed. Understanding habitat utilization patterns of larger, older and more fecund lionfish might shed light on ways to maximize the efficiency of lionfish culling programs. In this study, we analyzed data from ten locations in the Mexican Caribbean to determine the pattern between lionfish size, age, and the depth at which they live, and explored two possible explanations for the observed patterns: ontogenetic migration and harvesting effects. All locations signified a general presence of larger lionfish in deeper areas, confirming previous observations. Not enough evidence was found to suggest that the pattern is caused by culling. However, a likely explanation is ontogenetic migration, indicated by the presence of older lionfish in deeper waters. Results from this study highlight the importance of focusing management strategies on deeper waters to maximize effectiveness of lionfish removal programs."
output: 
  html_document:
      code_folding: hide
      toc_float: yes
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

suppressPackageStartupMessages({
  library(rnaturalearth)
  library(tidyverse)
  library(cowplot)
})
```

# Introduction
Invasive species are widely recognized as a major threat to the health and biodiversity of marine ecosystems. Yet, less than 16% of marine ecosystems are free of non-native species (Molnar et al. 2008). Non-native species tend to upset the balance of the ecosystem they invade by displacing the native populations and disrupting ecological processes, with associated costs of up to 120 billion USD annually (Pimentel et al. 2005). Lionfish, an invasive species originating in the Indo-Pacific and spreading throughout the western Atlantic, are no exception. Lionfish impact invaded areas by reducing native populations to levels of endangerment through predation and competition, affecting the health of coral reefs and of the overall ecosystem (Green et al. 2012). The effectiveness of lionfish management programs should be considered very important, with these programs being the main way of ensuring the minimization of lionfish impact in the future.

The success of these lionfish management programs lies in understanding habitat utilization patterns. However, current invasion management programs are limited by the depth at which divers are able to go to, thus, removal efforts focus solely on the shallow areas that lionfish populate. Previously, lionfish were believed to live only in shallow waters, but recently, observations of lionfish spotted in deeper areas have surfaced (Gress et al. 2017; Andradi-Brown et al. 2017). More importantly, anecdotal evidence suggests that larger and more fecund organisms live in deeper waters. However, no formal quantification of this pattern has been reported. This poses a major problem in the effectiveness of the programs, as larger lionfish remain safe from culling programs. As a result, larger lionfish living in deeper waters spread rapidly, facilitated by their lack of predators, opportunistic feeding habits, and fecundity (Gardner et al. 2015). With most lionfish management programs focusing mainly on controlling lionfish populations in shallower waters (< 30 m),  the refuge provided for lionfish by deeper waters is often overlooked in techniques regarding lionfish containment. Without more attention being focused on deeper waters, lionfish will continue to cause destruction within marine ecosystems. However, with research being done on habitat utilization of lionfish, management programs may be able to decrease the harm lionfish cause to marine ecosystems by focusing their efforts on deeper waters.

This study evaluates the relationship between lionfish size and the depth at which they live, and poses the hypotheses of ontogenetic migration and shallow-biased culling as possible explanations for the size-depth relationship. Analyzing this relationship before and after culling does not provide a comprehensive analysis, as the lengths of the lionfish are not exact, however, it is suggested by the data that ontogenetic migration may be one of the larger factors which are causing the pattern. Our focal analysis on the Mexican Caribbean may help us better understand how we can target lionfish and eliminate them from non-native areas.


# Methods

**Data Collection**

Our study took place at ten different sites throughout the Mexican Caribbean (Figure 1). We caught 109 lionfish using handheld nets and numbered containers, with no regard to age, size, depth, gender, or other factors. The depth (nearest 0.1 m) and exact location (latitude and longitude) at which they were found were immediately recorded upon capture, and their total length (TL to the nearest 1 mm) and total weight (TW nearest 0.01 g) were measured upon reaching the surface.

We converted TL to age with the von Bertalanffy growth equation:
$$A = t_0 + \frac{log(1-\frac {L}{L_{\infty}})}{-K}$$
Parameters ($L_{\infty} = 420$, $K = 0.88$, $t0 = 0.11$) for Yucatan were obtained from Johnson and Swenarton (2016).

```{r}
mex <- ne_countries(continent = c("north america"), scale = "medium", returnclass = "sf")
```

```{r, include = F}
ls <- ggplot(data = mex) +
  geom_sf() +
  lims(x = c(-110, -70), y = c(5, 35)) + #specify x and y limits
  geom_point(x = -87, # longitude of study area
             y = 20.35, # latitude of study area
             shape = 22, # square (?pch)
             size = 5, # square size
             color = "red" # square color
             ) +
  labs(x="Longitude", y="Latitude")
ls
```

```{r}
data_mexico <- read_csv(here::here("data", "RMP_lionfish.csv")) %>% 
  select(location, longitude, latitude) %>% 
  distinct()

data_mexico %>% 
  knitr::kable()
```

```{r}
coast <- sf::read_sf(dsn = here::here("data", "spatial", "coastline"), layer = "coastline")
```
 
```{r}
ss <- ggplot(data = coast) +
  geom_sf() +
  geom_point(data = data_mexico, mapping = aes(x = longitude, y = latitude), size = 1.7)+
  labs(x="Longitude", y="Latitude") +
  scale_x_continuous(breaks = seq(-87.5, -87.2, by = .1), limits = c(-87.35, -87.15)) +
  lims(y = c(20.375, 20.53))
```
 
```{r, fig.height=7, fig.width=7}
plot_grid(ls, ss, labels = "AUTO", ncol = 1)
```

**Data Analysis**

We evaluated the relationship that Depth had to lionfish TL, TW, and age with Pearson's correlation coefficient. Uppon determining the existance of positive correlations, we used linear regressions to fit a linear model relating these variables.

Additionally, we calculated batch fecundity ($f$; the total number of eggs produced as a function of TL) for each lionfish as $f = 308.67 \times TL -58265$ following Gardner et al., (2015). We grouped lionfish in 5-meter depth intervals, and estimated total egg production for each bin to show the invasion management implications of ontogenic habitat utilization.

While trying to analyze the relationship between size and depth before and after culling of lionfish in the Mexican Caribbean, it was determined that the post-culling data collected were insufficient to confirm a pattern of deepening caused by shallow-biased culling. Thus, more research should be done on the effect of culling on the habitat selection of lionfish, with attention being focused on including data from lower depths and larger lionfish to be accurate. All data analyses were performed in R Version 3.5.0, using ggplot2 for data visualization (R Core Team, 2018). The stargazer package in R (Hlavac, 2018) was used to create tables with regression coefficients (See Figure #).

# Results
```{r}
Linf = 420
T0 = 0.11
K = -0.88
library(tidyverse)
```

```{r}
fish <- read_csv(here::here("data", "RMP_lionfish.csv"), col_types = cols()) %>% 
  select(id, site, location, latitude, longitude, depth, temperature, total_length, total_weight) %>%
  mutate(age = T0 + (log(1-(total_length/Linf))/K), 
         logTL = log(total_length),
         logDepth = log(depth),
         logAge = log(age),
         logTW = log(total_weight)) %>%
  distinct()
```

```{r, fig.width = 6.5, fig.height = 12}
length_depth <- ggplot(data= fish, mapping = aes(x = total_length, y = depth)) +
  geom_point(mapping = aes(fill = location), size = 4, alpha = 0.8, shape = 21) +
  theme_bw() +
  labs(x="Total Length (mm)", y="Depth (m)", fill="Location") +
  geom_smooth(method = "lm") +
  theme(legend.position = "none")

weight_depth <- ggplot(data=fish, mapping = aes(x=log10(total_weight), y=depth)) +
  geom_point(mapping = aes(fill = location), size = 4, alpha = 0.7, shape = 21) +
  theme_bw() +
  labs(x=expression(log[10](Total~Weight~(gr))), y = "Depth (m)", fill = "Location") +
  geom_smooth(method = "lm") +
  theme(legend.position = c(0.2, 0.77),
        legend.background = element_blank()) +
  guides(fill = guide_legend(ncol = 2))

age_depth <- ggplot(data=fish, mapping = aes(x=age, y=depth)) +
  geom_point(mapping = aes(fill = location), size = 4, alpha = 0.7, shape = 21) +
  theme_bw() +
  labs(x="Age (years)", y = "Depth (m)", fill = "Location") +
  geom_smooth(method = "lm") +
  theme(legend.position = "none")

cowplot::plot_grid(length_depth, weight_depth, age_depth, ncol = 1, labels = "AUTO")
```

## Table
```{r, results = "asis"}
library(stargazer)
model1 <- lm(depth ~ total_length, data = fish)
model2 <- lm(depth ~ total_weight, data = fish)
model3 <- lm(depth ~ age, data = fish)
stargazer(model1, model2, model3, type = "html")
```


# Discussion and Conclusion
```{r}

depth_vec <- seq(0, 40, by = 5)

# Load data, mutate fecundity (f) and make depth into bins (depth_bin)
# then group by depth and calculate total egg production

fecundity_data <- read.csv(here::here("data", "RMP_lionfish.csv")) %>%
  mutate(f = -58265 + (308.67 * total_length),
         depth_bin = findInterval(depth, depth_vec)) %>%
  group_by(depth_bin) %>%
  summarize(total_f = sum(f))

# Plot fecundity at depth
ggplot(data = fecundity_data, mapping = aes(x = depth_bin, y = total_f), size = 4, alpha = 0.8, shape = 21) +
  geom_point() +
  theme_bw() +
  labs(x="Depth", y="Fecundity") +
  ggtitle("Lionfish Depth-Fecundity Relationship") +
  geom_smooth(method = "lm")
```


# References
Accadia, T., Acernese, F., Alshourbagy, M., Amico, P., Zhang, Z., et al. 2012. Virgo: a laser interferometer to detect gravitational waves. Journal of Instrumentation 7(03), pp. P03012–P03012.

Andradi-Brown, D.A., Vermeij, M.J.A., Slattery, M., Lesser, M., Bejarano, I., Appeldoorn, R., Goodbody-Gringley, G., Chequer, A.D., Pitt, J.M., Eddy, C., Smith, S.R., Brokovich, E., Pinheiro, H.T., Jessup, M.E., Shepherd, B., Rocha, L.A., Curtis-Quick, J., Eyal, G., Noyes, T.J., Rogers, A.D. and Exton, D.A. 2017. Large-scale invasion of western Atlantic mesophotic reefs by lionfish potentially undermines culling-based management. Biological invasions 19(3), pp. 939–954.

Gardner, P. G., Frazer, T. K., Jacoby, C. A., & Yanong, R. P. (2015). Reproductive biology of invasive lionfish (Pterois spp.). Frontiers in Marine Science, 2, 7.

Green, S.J., Akins, J.L., Maljković, A. and Côté, I.M. 2012. Invasive lionfish drive Atlantic coral reef fish declines. Plos One 7(3), p. e32596.  

Gress, E., Andradi-Brown, D.A., Woodall, L., Schofield, P.J., Stanley, K. and Rogers, A.D. 2017. Lionfish (Pterois spp.) invade the upper-bathyal zone in the western Atlantic. PeerJ 5, p. e3683.  

Hlavac, Marek (2018). stargazer: Well-Formatted Regression and Summary Statistics Tables.
 R package version 5.2.2. https://CRAN.R-project.org/package=stargazer

Johnson, E.G. and Swenarton, M.K. 2016. Age, growth and population structure of invasive lionfish (Pterois volitans/miles) in northeast Florida using a length-based, age-structured population model. PeerJ 4, p. e2730.  

Molnar, J.L., Gamboa, R.L., Revenga, C. and Spalding, M.D. 2008. Assessing the global threat of invasive species to marine biodiversity. Frontiers in Ecology and the Environment 6(9), pp. 485–492.  

Pimentel, D., Zuniga, R. and Morrison, D. 2005. Update on the environmental and economic costs associated with alien-invasive species in the United States. Ecological Economics 52(3), pp. 273–288.  

R Core Team (2018). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

